---
title: "\\textbf{GLM}"
subtitle: "\\textbf{Data Analysis Assignment 03 - Multinomial GLM}"
format:
  pdf:
    include-in-header:
      text: |
        \addtokomafont{disposition}{\rmfamily}
editor: visual
geometry:
     - top=10mm
     - left=12mm
     - right=12mm
header-includes:
   - "\\usepackage{booktabs}"
   - "\\usepackage{titling}"
author: "Divya Sharma (ds655)"
---

# Overview

*GLMs* (Generalized Linear Models) are a flexible extension of linear regression models. They allow the response variable to have a distribution other than a normal distribution, such as binomial, Poisson, etc.

Like linear regression, GLMs relate the response variable to explanatory variables using a linear function. But they also include a *link function* that transforms the response variable so its distribution matches the assumptions.

The key components of a GLM are:

-   The response variable Y and its distribution (e.g. binomial, Poisson)

-   The linear predictor - a linear function of the explanatory variables (like in linear regression)

-   The link function g() that transforms the expected value of Y to match the linear predictor

The link function connects the linear predictor to the expected response, allowing a non-normal distribution for Y.

For example, a GLM could model student test scores based on hours studied and previous GPA. The response variable test score may follow a Poisson distribution. The linear predictor combines the explanatory variables hours studied and GPA. The log link function transforms the expected test score to match this linear combination.

A multinomial GLM handles a categorical response variable with more than 2 categories. The purpose is to model the probability of different outcomes. For example, a multinomial GLM could predict political party affiliation based on age, income, gender, etc. The response is party, a categorical variable with multiple outcomes. The linear predictor combines the explanatory variables. The logit link connects this to the probability of each party.

Some potential research questions for a GLM model can be:

-   How does income level correlate with Republican vs Democrat affiliation?

-   Do gender and age interact in predicting party?

-   What variables most strongly predict party affiliation?

In summary, a multinomial GLM can model a categorical response to explain the factors influencing different outcomes. The link function handles the non-normal distribution.

# Probability Distribution

Briefly describe the probability distribution that is assumed for the outcome. What is the support? What are the parameters and what values can they take?

# Model

Write out the general form of your GLM. What is the link function and why is it appropriate for that type of outcome? What are the model assumptions?

# Data Example

-   Introduce the dataset. Provide a few summary statistics and/or plots. Include the fact that this is a simulated dataset.

-   Fit the model. Include all relevant code, including `library()` for packages if needed.

-   Explain how to interpret coefficient estimates for the predictors.

-   Show and describe a plot that illustrates the results of the model.

-   Describe how to assess the model and include the code to do so. Include how to assess any assumptions that are unique to that model (e.g., proportional odds, overdispersion). *Note: For the ordinal model, assess the assumption using the method shown in the class exercise, not the hypothesis test shown in the videos.*

```{r,include=FALSE}
#Importing Packages
library(dplyr, quietly = T)

```

```{r,include=FALSE}
#Reading Base Data
base_data <- read.csv("https://github.com/DivyaSharma0795/IDS702_Data_Analysis_Assignment_03/raw/main/Resources/DA3_Multinomial.csv")
```

```{r}
base_data
```
